The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-09-21 manuscript/rebuttal.Rmd

### R2 Point 1

> 1. Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case
> studies or practical examples that demonstrate the effectiveness of your recommendations.
> Highlighting specific instances where these best practices have enhanced reproducibility or solved
> particular issues could substantiate your points more effectively.

RESPONSE: To demonstrate the importance of computational reproducibility, we elaborate on the
"forensic bioinformatics" case study of Baggerly and Coombes (2010) @Baggerly2010-nr critique of Potti
et al (2006) @Potti2006-yk in the introduction:

> For example, in 2006 an article entitled "Genomic signatures to guide the use of chemotherapeutics"
> generated a great deal of interest as it was an early application of high throughput transcriptomics
> in the prediction of individual patient responses to different chemotherapies @Potti2006-yk.
> After observing some unusual features of the patient group, Baggerly and Coombes attempted
> reproduction of some of the key findings @Baggerly2010-nr.
> Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
> techniques to piece together how the dataset was originally analyzed.
> Their investigations found a litany of issues.
> Firstly, the labeling of patients in the test set as "responders" or "non-responders" had been reversed
> in the original analysis.
> Secondly, some of the patients were included more than once (some up to four times) in the analysis,
> likely to cause major distortions in results.
> Confusingly, some of the reused data has inconsistent grouping; ie: some of the samples are labeled
> both sensitive and resistant.
> Additional errors include two cases where results (charts) were ascribed to the wrong drug.
> Baggerly and Coombes highlight that such mistakes can inadvertently occur when conducting unscripted
> data analysis such as using spreadsheets, and such problems can be obscured by a lack of documentation.
> The article underwent two corrigenda, but was ultimately retracted in 2011, as the authors were not
> able to reproduce the findings themselves due to "corruption of several validation data sets"
> @Potti2011-ut.
> As the array findings were the basis for clinical trials where patients were allocated to treatments,
> the flawed data analysis may have harmed patients given the wrong drug in the
> period 2007-2010.
> In 2010, Duke University terminated the trials and suspended the lead author, Dr Anil Potti, who
> later resigned.
> Duke was served eight lawsuits by families of affected patients seeking compensation for exposure to
> harmful and unnecessary chemotherapy, which were settled out of court @Dyer2015-my.
> This worst-case scenario emphasizes that computational reproducibility is crucial for
> translating bioinformatics research into real-world outcomes.

@@

### R2 Point 5

> 5. Engaging Conclusion: Reworking the conclusion to be more impactful could be beneficial. Instead of
> merely summarizing the recommendations, it would be useful to indicate the potential trajectory of
> computational research if these recommendations are adopted, as well as highlighting the potential
> risks if they aren't.

Response: We have completely rewritten the conclusion.
Instead of reiterating the five pillars, we provide the following:

> As the chemotherapeutics case study highlights, there are significant risks to poorly conducted
> bioinformatics research, and current standards are falling short.
> If adopted widely, the transparency and reproducibility measures proposed here will reduce the chance of
> such disasters from happening again.
> Researchers, their institutions, publishers and funders each have a major role to play in fighting the
> reproducibility crisis by encouraging highly reproducible research practices.

@@

## Reviewer 3

### R3 Point 1

> The paper is mostly well written and addresses an important topic: computational reproducibility and
> tools to achieve it. The authors describe five "pillars" under which various tools and techniques can
> be categorized. The paper is pretty comprehensive in what it covers and has useful supplementary
> material. My main issue with the paper is that it reiterates many things that have been covered in
> other papers. The authors cite 11 such papers. Many of the topics covered in submitted paper have
> already been covered well in other papers. I am most familiar with reference 13, which covers many of
> the same topics, although this paper goes into more detail on many issues. The authors could do more
> to differentiate their paper from previous ones and perhaps remove some topics that are already
> covered well elsewhere.

RESPONSE: We thank reviewer 3 for their careful assessment of our article and appreciate the suggestions
which have improved the manuscript.
Indeed, this article does reiterate a great deal of what has already been said, it is a review after
all.
We note that each of those recommendation articles mentioned have different emphases.
For example, some of them don't speak to the benefits of literate programming,
while others do not mention the importance of documentation.
Some are focused on specific concepts like infrastructure, containerization or
continuous analysis.
Some are proof of principle workflows, which present a rigid combination of tools.
Here, we describe the ways computational researchers can make their own work more reproducible.
We are assuming that the readers are not advanced analysts, so some topics like pipelining systems
are covered in less detail and we refer the reader to those relevant reviews.
What does make this article unique is the five pillars infographic, and specific combination of
recommendations that incorporates new material (eg: Software Heritage, Guix, Continuous Analysis).
This is useful, because bioinformatics analysts can use the infographic to benchmark their projects
and identify areas where changes are needed.
Our article also provides a handy list of recommended tutorials as a supplementary resource.

@@

> 9. How do you get from a notebook to an actual paper submission if you have to do custom formatting
> of the document, including references?
> My understanding is that this is still not possible, but please correct me if I'm wrong.

RESPONSE: It is possible to achieve accurate journal-specific formatting of references with a Rmarkdown
document.
It requires a few things:

* A .bib file that contains the bibliography in bibtex format.

* A .csl file that specifies the citation style.
These are available from [GitHub](https://github.com/citation-style-language/styles).

@@

RESPONSE: Citations have been added for Podman and Apptainer/Singularity.
Rkt was removed, as it appears that development of the project has ceased.

### R3 Point 16

> 16. Page 9, lines 7-8: I believe you, but I am not aware of evidence that supports this claim.

RESPONSE: The claim made in the manuscript is that "Docker remains the most widely used
containerization system for web development and research".
We do not have exact usage data on this, however we can look at other metrics like GitHub stars.

@@

Singularity: 3.1k

(25th August).

Based on this we have altered the text:

> While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
> [@Kurtzer2017-zq], Docker remains the most widely used containerization system (according to GitHub
> stars).

@@

> 5. Use file formats that are machine-readable and compatible with many different types of software.
> Some examples include comma- and tab-separated values (CSV/TSV) formats, eXtensible Markup Language (XML),
> JavaScript Object Notation (JSON), Hierarchical Data Format version 5 (HDF5) and Apache Parquet.

### R3 Point 25

> 25. One thing that could be added is something about Common Workflow Language.
It's a community-supported specification for accomplishing many of the objectives described here.
There are some recent papers about this.

RESPONSE: Thanks for raising this. We have added CWL to the section mentioning pipeline solutions, and
referred readers to dedicated reviews on the topic.

### 2023-09-21 manuscript/rebuttal2.Rmd

---
title: "Response to reviewer comments, part II (The five pillars)"
author: "Mark Ziemann, Pierre Poulain, Anusuiya Bora"
date: "`r Sys.Date()`"
output: word_document
code_folding: hide
fig_width: 7
fig_height: 6
bibliography: 5pillars.bib
csl: plos-computational-biology.csl
---

Dear Editor,

Again, many thanks for your consideration of our manuscript.
Below is our point-by-point response to Reviewer 3's remining comments.

### Point 2

> In my view, the paper is not suitable for "anyone who analyzes biological data." After all, the
> journal is Briefings in Bioinformatics.
> Much of the material is too technical for the stated audience.
> The authors have provided "Practical guides for implementing ‘Five Pillar’ principles."
> This may be helpful to some, but it is not realistic to assume that people will consume all of
> those materials before reading the paper or that by doing so they will be able to understand all of
> the technical details in the manuscript.
> In my view, the intended audience should be clarified in the manuscript.

RESPONSE: Indeed much of the manuscript is oriented towards gold-standard practices and this could be
out of the immediate reach for many biological data analysts, but this is not a futile exercise.
Firstly, papers such as this one set out a road map of self-education which may take years to
accomplish.
Secondly, where self-education reaches its limits, they can seek workshops and other training to
strengthen these priority areas.

Therefore, we have altered the phrase to recognise the journey of bioinformaticians in training.

> The intended audience is bioinformatics data analysts and bioinformaticians in training, however the
> principles described here could equally apply to other domains of study.

If the reviewer has a specific suggestion, then it would be helpful to provide this.

### Point 10

> 10. What about when your data files are too large to fit on a personal computer? Please address in
the manuscript.

RESPONSE: In accordance with this suggestion we have added the following passage to the literate
programming section.

> While it is common to run literate programs on a PC with graphical interfaces such as Rstudio,
> they can also be executed in "headless" mode on high performance computing clusters or cloud
> servers using command-line interfaces to take advantage of greater computational resources.
> Headless mode is important for integrating literate scripts into larger workflows and for
> automated testing.

### Point 17

> 17. The Biocontainers project should be mentioned.

RESPONSE: We have added an additional mention of BioContainers which appears just after where DockerHub
is mentioned.

> In bioinformatics, containers are already extensively used.
> For example, BioContainers is a registry for sharing bioinformatics software containers
> [@Da_Veiga_Leprevost2017-yw].
> Containers are equally useful to data analysts by encapsulating the environment in which analytical
> scripts are executed.

### Point 20

> 20. Figure 5: I don't think it's really necessary to resummarize the FAIR principles.
> People can go to the source article for that.
> I don't mean to be snarky, but everyone has limited time. I still believe that spelling out these
> points only serves to make the manuscript longer than necessary and thus make it less likely for
> people with "limited time" to read it.
> FAIRness is not the main thrust of this manuscript.

RESPONSE: We appreciate the reviewer's point of view.
We have removed mention of FAIR and we refer readers to the literature on this topic.
FAIR has also been removed from the pillars infographic and Figure 5 is also gone.

### R3 Point 21

> 21. Page 11, line 13: I disagree that repositories like GEO and SRA are FAIR.
There are lots of problems with FAIRness in these repositories.

RESPONSE: Possibly the reviewer hasn't seen the rebuttal letter had it's own bibliography.
Anyway, since "FAIR" is not described anymore, it simplifies the description of these databases,
which satify most best practices. We cite the work of Bonaretti et al 2019.
We also cite Hassenruck et al 2021, who report deficiencies in these databases, but spread the blame
between database administrators, researchers and other stakeholders.
The two important passages we have amended are provided below:

> In the wake of the Human Genome Project, a number of repositories for specific biological data types
> were established including GEO, SRA, ENA and PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Martens2005-wo].
> Although these repositories suffer issues around interoperability and reusability, they do support
> data reuse through findability and accessibility [@Bonaretti2019-zs; @Hassenruck2021-mi].

...

> Researchers should be aware, however, that these repositories are often not moderated,
> and that it is their responsibility to provide sufficient metadata.

## Bibliography

### 2023-09-21 guides/practical_guides.Rmd
documentation edits
2nd round comments

Mark Ziemann<sup>1*</sup>, Pierre Poulain<sup>2</sup>,
Anusuiya Bora<sup>1</sup>

Author information

1. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.

2. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.

Correspondance: m.ziemann@deakin.edu.au

| Author | ORCID |
| --- | --- |
| Mark Ziemann | 0000-0002-7688-6974 |
| Pierre Poulain | 0000-0003-4177-3619 |
| Anusuiya Bora | 0009-0006-2908-1352 |

</details>
</p>

@@

The five pillar approach is a synthesis of over a decade of learnings around computational
reproducibility, and contains six key recommendaions and many best practices.
These may be overwhelming for any data scientists, especially beginners and novices,
so here we have tried to put together some recommended resources to enable practitioners to put these
principles into practice.
One of the questions practitioners may have is *"Where do I start?"* and
*"In what order should I learn/master these principles?"*.
The following sections have been arranged in order, and have been submitted to the
[Internet Archive](https://archive.org/) for posterity.

@@

### Introduction to Unix and Shell programming

@@

@@

@@

@@

## Moving from "dev" to "prod": testing and continuous analysis

Continuous analysis involves automatic execution and testing of code, which is important when changes
are made to code or data.
Results generated in HTML are automatically updated.
Badges on the repository indicate the status of the codebase in terms of completing with or without
errors.

@@

>>>>

@@

The five pillar approach is a synthesis of over a decade of learning around computational
reproducibility, and contains seven key recommendations and many best practices.
These may be overwhelming for any data scientists, especially beginners and novices,
so here we have tried to put together some recommended resources to enable practitioners to put these
principles into practice.
One of the questions practitioners may have is *"Where do I start?"* and
*"In what order should I learn/master these principles?"*.
The following sections have been arranged in order, and have been submitted to the
[Internet Archive](https://archive.org/) for posterity.

@@

### Introduction to Unix and shell programming

@@

### Introduction to Galaxy: bioinformatics in the browser

* [Galaxy: A very short introduction](https://www.youtube.com/watch?v=VZoz3k5EehI&ab_channel=GalaxyProject)

@@

* [Tutorial: Write your next paper in MyST Markdown with data, code & Jupyter notebooks](https://www.youtube.com/watch?v=1BWYLzQd8Ds&t=4s&ab_channel=curvenote)

@@

* [Guide: A Quick Guide to Organizing Computational Biology Projects](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1000424)

@@

## Continuous analysis/continuous validation

Continuous analysis involves automatic execution and testing of code, which is important when changes
are made to code or data.
Results generated in HTML are automatically updated.
Badges on the repository indicate the status of the codebase in terms of completing with or without
errors.
Transparency and reproducibility help validation which can be continously maintained.

@@

* [Tool: Drone: System for continuous delivery](https://www.drone.io/)

### 2023-09-21 manuscript/5pillars.Rmd
2nd round comments

* Correspondence: m.ziemann@deakin.edu.au

| `Author      ` |  `ORCID             ` |
| --- | --- |
| Mark Ziemann | 0000-0002-7688-6974 |
| Pierre Poulain | 0000-0003-4177-3619 |
| Anusuiya Bora | 0009-0006-2908-1352 |

@@

Computational reproducibility is a simple premise in theory, but is difficult to
achieve in practice.
Building upon past efforts and proposals to maximize reproducibility and rigor in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) literate programming, (2) code version control and sharing,
(3) compute environment control, (4) FAIR and persistent data sharing and (5) documentation.
These practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatic data analysts, but should be relevant for other
domains.

@@

A number of guides recommending enhanced computational reproducibility practices have been developed
[ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
@Stodden2016-aj; @Wilson2017-dv; @Beaulieu-Jones2017-vj; @Gruning2018-sg; @Kim2018-nv; @Akalin2018-eg;
@Brito2020-ld; @Peikert2021-eq; @Rodrigues2023-fw; @DuPre2022-bk; @Krafczyk2021-xv].
Inspired by these principles, we presents the *five pillars of reproducible computational research*
(**Figure 1**).
Here, the emphasis is on practical reproducibility with an increased focus on programming practices,
transparent reporting and the role of computational environments.
The intended audience is anyone who analyzes biological data (not tool developers), however the
principles described here could equally apply to other domains of study.
There are clear parallels with the established three pillars of the open science framework,
(open data, code and papers) [@Gorgolewski2016-hv].

@@

labs <- c("Literate\nprogramming",
  "Code version\ncontrol &\npersistent\n sharing",
  "Compute \nenvironment \ncontrol",
  "FAIR &\npersistent\ndata sharing",
  "Documentation")

@@

Literate programming also works neatly with the concept of the "executable paper", the idea that
the data analysis underlying an entire study can be reproduced with one or a few commands
[@Strijkers2011-py ; @Editorial2017-yt; @Lasser2020-mm].
A typical genomics/bioinformatics study could involve one large literate script, or be broken down into
smaller scripts, where each one contains the code for generating a part of the article.
A multi-script approach would benefit from a master script that executes each component.
This further makes the job of reproducibility easier.

@@

In bioinformatics, sharing code is becoming standard practice for reproducibility and transparency,
and is a requirement for many specialized journals [@Cadwallader2022-dq].
Code sharing appears to improve the rate of article citations [@Maitner2023-pv].
One of the most popular ways to share research code is through online software repositories
[@Cadwallader2022-od] with integrated version control.
A version control system (sometimes called "source control") is a type of program that "tracks changes"
made to sets of files, typically other computer program source code, scripts and documentation.
A set of files under version control is called a "repository" and represents a project or sub-project.
Version control is used extensively by software developers and is considered one of the key best
practices in software engineering.
Distributed version control systems (DVCS) involve a central web-accessible server hosting a
repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
Having a central, publicly available node assists in disseminating changes within teams and releasing
code to consumers.
There are many such DVCSs (Subversion, Git, mercurial, etc), but `git` has emerged as the most popular
solution due to its many powerful features, speed/efficiency and large community and ecosystem.

@@

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerized applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerization system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
[@Kurtzer2017-zq], Docker remains the most widely used containerization system (according to GitHub
stars as of August 2023).
Docker has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].

@@

### FAIR and persistent data sharing

@@

The field of genomics has a long history of data sharing @Brazma2001-yk, which is supported by
policy leadership by funding bodies
[@National_Health_and_Medical_Research_Council2019-yx;
@Office_of_The_Director_National_Institutes_of_Health2023-ss] and data sharing policies of journals
(eg [@Packer2018-ru; @Hanson2011-ub ; @Cheifet2020-wu]), although there are some challenges.
Some researchers would rather keep data for their private use.
Some may keep data private due to concerns over participant privacy.

While deidentification of data has been widely used to safeguard participant privacy, this may not
be suitable for genomics data due to the availability of existing public genetic data sets that heighten
reidentification risks @Heeney2011-ri.
For example, the surnames of some male participants can be inferred based on the Y chromosome variants
seen in deidentified genomic data together with public geneology web sites @Gymrek2013-fz.
To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
frameworks designed to protect personal data in a way that preserves reproducibility
[@Rehm2021-kt;@Thorogood2021-kt].

In order to maximize the value of shared data for reuse and reproducibility, it needs to be Findable,
Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
These principles are summarized in **Figure 5**.

![Figure 5. Overview of the FAIR data principles @Kalendralis2021-jb.](../img/gr2.jpg)

High throughput biology has already built a number of data repositories for specific data types mostly
conforming to FAIR principles.
Repositories like GEO, SRA, ENA and PRIDE were developed in the aftermath of the draft human
genome assembly and have become well known to omics researchers
[@Edgar2002-hs; @Kodama2012-bn; @Martens2005-wo].
The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
FAIR repositories that accepts data from other domains of study like ecology,
physiology, molecular simulation, social sciences and computing [@Pampel2013-ep].
If no specialized repository exists, then a general-purpose repository such as
[Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
should be considered. Researchers should be aware, however, that these repositories are often not moderated, 
and that it is their responsibility to make their data FAIR, in particular by providing sufficient metadata.
To this aim, general recommendations have been proposed to enhance data FAIRness in ecology, but which
can also be applied to other disciplines [@Roche2015-aq].

@@

In addition, the `README` should outline the contents of the repository, how users can contribute to
the project and how to report issues, such as code errors and discrepancies.
The `README` should also list the contributors to the codebase and acknowledge any financial support.
It may also include a link to the research article, citation instructions and a mention of the software
license.
A recently described set of recommendations for making software FAIR describes further measures
that data analysts could use to enhance their code documentation @Patel2023-qr.

>>>>

(*) Correspondence: m.ziemann@deakin.edu.au

| Author | ORCID |
| --- | --- |
| Mark Ziemann | 0000-0002-7688-6974 |
| Pierre Poulain | 0000-0003-4177-3619 |
| Anusuiya Bora | 0009-0006-2908-1352 |

@@

Computational reproducibility is a simple premise in theory, but is difficult to
achieve in practice.
Building upon past efforts and proposals to maximize reproducibility and rigor in
bioinformatics, we present a framework called the five pillars of reproducible computational
research.
These include (1) literate programming, (2) code version control and sharing,
(3) compute environment control, (4) persistent data sharing and (5) documentation.
These practices will ensure that computational research work can be
reproduced quickly and easily, long into the future.
This guide is designed for bioinformatic data analysts, but should be relevant for other
domains.

@@

A number of guides recommending enhanced computational reproducibility practices have been developed
[ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
@Stodden2016-aj; @Wilson2017-dv; @Beaulieu-Jones2017-vj; @Gruning2018-sg; @Kim2018-nv; @Akalin2018-eg;
@Brito2020-ld; @Peikert2021-eq; @Rodrigues2023-fw; @DuPre2022-bk; @Krafczyk2021-xv].
Inspired by these principles, we presents the *five pillars of reproducible computational research*
(**Figure 1**).
Here, the emphasis is on practical reproducibility with an increased focus on programming practices,
transparent reporting and the role of computational environments.
The intended audience is bioinformatics data analysts and bioinformaticians in training, however the
principles described here could equally apply to other domains of study.
There are clear parallels with the established three pillars of the open science framework,
(open data, code and papers) [@Gorgolewski2016-hv].

@@

labs <- c("Literate\nprogramming",
  "Code version\ncontrol &\npersistent\n sharing",
  "Compute \nenvironment \ncontrol",
  "Persistent\ndata sharing",
  "Documentation")

@@

Literate programming also works neatly with the concept of the "executable paper", the idea that
the data analysis underlying an entire study can be reproduced with one or a few commands
[@Strijkers2011-py ; @Editorial2017-yt; @Lasser2020-mm].
A typical genomics/bioinformatics study could involve one large literate script, or be broken down intosmaller scripts, where each one contains the code for generating a part of the article.
A multi-script approach would benefit from a master script that executes each component.
This further makes the job of reproducibility easier.

While it is common to run literate programs on a PC with graphical interfaces such as Rstudio,
they can also be executed in "headless" mode on high performance computing clusters or cloud
servers using command-line interfaces to take advantage of greater computational resources.
Headless mode is important for integrating literate scripts into larger workflows and for
automated testing.

@@

In bioinformatics, sharing code is becoming standard practice for reproducibility and transparency,
and is a requirement for many specialized journals [@Cadwallader2022-dq].
Code sharing appears to improve the rate of article citations [@Maitner2023-pv].
One of the most popular ways to share research code is through online software repositories
with integrated version control [@Cadwallader2022-od].
A version control system (sometimes called "source control") is a type of program that "tracks changes"
made to sets of files, typically other computer program source code, scripts and documentation.
A repository is simply a set of files under version control that represents a project or sub-project.
Version control is used extensively by software developers and is considered one of the key best
practices in software engineering.
Distributed version control systems (DVCS) involve a central web-accessible server hosting a
repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
Having a central, publicly available node assists in disseminating changes within teams and releasing
code to consumers.
There are many such DVCSs (Subversion, Git, mercurial, etc), but `git` has emerged as the most popular
solution due to its many powerful features, speed/efficiency and large community and ecosystem.

@@

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerized applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerization system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
[@Kurtzer2017-zq], Docker remains the most widely used containerization system (according to GitHub
stars as of August 2023).
Docker has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].
In bioinformatics, containers are already extensively used.
For example, BioContainers is a registry for sharing bioinformatics software containers
[@Da_Veiga_Leprevost2017-yw].
Containers are equally useful to data analysts by encapsulating the environment in which analytical
scripts are executed.

@@

### Persistent data sharing

@@

The field of genomics has a long history of data sharing @Brazma2001-yk, which is supported by
policy leadership by funding bodies
[@National_Health_and_Medical_Research_Council2019-yx;
@Office_of_The_Director_National_Institutes_of_Health2023-ss] and data sharing policies of journals
(eg [@Packer2018-ru; @Hanson2011-ub ; @Cheifet2020-wu]).
Best practices for research data sharing have been proposed [@Wilkinson2016-qq], as have
recommendations around sharing human genetic/genomic data [@Byrd2020-ri].
While deidentification of data has long been used to safeguard participant privacy, this may not
be suitable for genomics data due to the availability of existing public genetic data sets that heighten
reidentification risks @Heeney2011-ri.
For example, the surnames of some male participants can be inferred based on the Y chromosome variants
seen in deidentified genomic data together with public geneology web sites @Gymrek2013-fz.
To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
frameworks designed to protect personal data in a way that preserves reproducibility
[@Rehm2021-kt; @Thorogood2021-kt].

In the wake of the Human Genome Project, a number of repositories for specific biological data types
were established including GEO, SRA, ENA and PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Martens2005-wo].
Although these repositories suffer issues around interoperability and reusability, they do support
data reuse through findability and accessibility [@Bonaretti2019-zs; @Hassenruck2021-mi].
The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
repositories that accepts data from other domains of study like ecology,
physiology, molecular simulation, social sciences and computing [@Pampel2013-ep].
If no specialized repository exists, then a general-purpose repository such as
[Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
should be considered.
Researchers should be aware, however, that these repositories are often not moderated,
and that it is their responsibility to provide sufficient metadata.
To this aim, general recommendations have been proposed to enhance data reuse in ecology, but which
can also be applied to other disciplines [@Roche2015-aq].

@@

In addition, the `README` should outline the contents of the repository, how users can contribute to
the project and how to report issues, such as code errors and discrepancies.
The `README` should also list the contributors to the codebase and acknowledge any financial support.
It may also include a link to the research article, citation instructions and a mention of the software
license.
A recently described set of recommendations for documenting software describes further measures
that data analysts could use to enhance their code documentation @Patel2023-qr.
